The effects of drought on land cover change and vegetation productivity in continental Chile

### 2024-05-09 article/paper_drought_Chile2023.tex
update version may 2024

\usepackage{lineno}\linenumbers \usepackage{multirow} \usepackage{lscape} \newcommand{\blandscape}{\begin{landscape}} \newcommand{\elandscape}{\end{landscape}}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{Journal Name}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The effects of drought on land cover change and vegetation productivity in continental Chile},
  pdfauthor={Francisco Zambrano; Anton Vrieling; Francisco Meza; Iongel Duran-Llacer; Francisco Fernández; Alejandro Venegas-González; Nicolas Raab; Dylan Craven},
  pdfkeywords={drought, land cover change, vegetation
productivity, ecosystem},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\author[8,9]{Francisco Fernández%
%
}
 \ead{francisco.fernandez@uss.cl} 
\author[]{Alejandro Venegas-González%
%
}

@@

\author[11,12]{Dylan Craven%
%
}

\affiliation[1]{organization={Hémera Centro de Observación de la Tierra,
Facultad de Ciencias, Escuela de Ingeniería en Medio Ambiente y
Sustentabilidad, Universidad Mayor},city={Santiago},country={7500994,
Chile},countrysep={,},postcodesep={}}
\affiliation[2]{organization={Observatorio de Sequía para la Agricultura
y la Biodiversidad de Chile (ODES), Universidad
Mayor},city={Santiago},country={7500994,
Chile},countrysep={,},postcodesep={}}
\affiliation[3]{organization={Faculty of Geo-Information Science and
Earth, University of Twente},city={Enschede, The
Netherlands},postcodesep={}}
\affiliation[4]{organization={Facultad de Agronomía y Sistemas
Naturales, Pontificia Universidad Católica de
Chile.},city={Santiago},country={Chile},countrysep={,},postcodesep={}}
\affiliation[5]{organization={Instituto para el Desarrollo Sustentable.
Pontificia Universidad Católica de
Chile},city={Santiago},country={Chile},countrysep={,},postcodesep={}}
\affiliation[6]{organization={Centro Interdisciplinario de Cambio
Global, Pontificia Universidad Católica de Chile, Santiago,
Chile.},,postcodesep={}}
\affiliation[7]{organization={Hémera Centro de Observación de la Tierra,
Facultad de Ciencias, Universidad
Mayor,},city={Santiago},country={7500994,
Chile},countrysep={,},postcodesep={}}
\affiliation[8]{organization={Center of Economics for Sustainable
Development (CEDES), Faculty of Economics and Government, Universidad
San
Sebastian},city={Santiago},country={Chile},countrysep={,},postcodesep={}}
\affiliation[9]{organization={Center of Applied Ecology and
Sustainability
(CAPES)},city={Santiago},country={Chile},countrysep={,},postcodesep={}}
\affiliation[10]{organization={},,postcodesep={}}
\affiliation[11]{organization={, GEMA Center for Genomics, Ecology \&
Environment, Universidad Mayor, Camino La Pirámide Huechuraba 5750,
Santiago, Chile},,postcodesep={}}
\affiliation[12]{organization={Data Observatory Foundation, Santiago,
Chile},,postcodesep={}}

@@

\begin{abstract}
Chile has experienced a persistent decrease in water supply, which
impacts the hydrological system and vegetation development. This
persistent period of water scarcity has been defined as a mega-drought.
There are few studies about the relationship between drought and
ecosystem changes that can help with a better understanding of
ecological drought. The aim of our study is to evaluate the interaction
of drought, land cover change, and vegetation productivity over
continental Chile. To assess drought, we used drought indices for
atmospheric evaporative demand (AED), water supply, and soil moisture
from short- (1, 3, 6 months) to long-term (12, 24, 36 months) time
scales. We derived the drought indices using monthly ERA5-Land
reanalysis data from 1981 to 2023. We used the Moderate-Resolution
Imaging Spectroradiometer (MODIS) datasets to derive information on
annual land cover and monthly vegetation productivity. Our results
showed that except for the Austral part, Chile has a temporal decreasing
trend in water supply, and across the whole country, there is an
increase in AED. These trends become stronger over longer time scales.
We found a negative trend in vegetation productivity in the
north-central area, which is more prominent for shrubland and savanna as
compared to croplands and forests. The anomaly in soil moisture over the
past 12 months (SSI-12) is the most important variable explaining these
changes, followed by anomalies in accumulated precipitation over one to
two years (SPI-12 and SPI-24). The variable importance obtained by
random forest models indicates that drought is explaining about 20--30\%
of the change in land cover surface across Chile for forest, grassland,
shrubland, and savanna but has no relation to the changes in croplands.
The increase in AED is the main variable associated with the change in
land cover, followed by a reduction in precipitation and soil moisture.
Our findings provide insightful information that could assist in
developing adaptation measures for Chilean ecosystems to cope with
climate change and drought. Also, this study could contribute to a
better comprehension of ecological drought.
\end{abstract}

\begin{keyword}
    drought \sep land cover change \sep vegetation productivity \sep 
    ecosystem
\end{keyword}
\end{frontmatter}
    \captionsetup{justification=raggedright,singlelinecheck=false}

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, boxrule=0pt, sharp corners, breakable]}{\end{tcolorbox}}\fi

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Drought is often classified as 1) meteorological when precipitation in a
specific period remains below the mean precipitation experienced in the
same period during multiple years (more than 30 years usually), 2)
hydrological when these anomalies last for long periods (months to
years) and affect water systems, and 3) agricultural when the deficit
negatively impacts plant health and leads to decreased productivity of
crops or pastures \citep{Wilhite1985}. However, because drought is also
influenced by human activities, \citet{Loon2016} and
\citet{AghaKouchak2021} expanded the drought definition for the
Anthropocene, indicating that the feedback of human decisions and
activities should also be considered (i.e., anthropogenic drought).
Droughts lead to increased tree mortality \citep{Cheng2024} and induces
alterations in land cover and land use, ultimately affecting ecosystems
\citep{Crausbay2017}. Even though many ecological studies have at times
mistakenly considered ``dry'' conditions as ``drought''
\citep{Slette2019}. Ecological drought can be defined as \emph{``an
episodic deficit in water availability that drives ecosystems beyond
thresholds of vulnerability, impacts ecosystem services, and triggers
feedback in natural and/or human systems''} \citep{Crausbay2017}. In
light of current global warming, it is crucial to study the interaction
between drought and ecosystems in order to understand their feedback and
impact on future water security \citep{Bakker2012}.

Global warming, as a result of human-induced greenhouse gas emissions,
has increased the frequency and intensity of drought, according to the
sixth assessment report (AR6) of the Intergovernmental Panel on Climate
Change (IPCC) \citep{IPCC2023}. The evidence supporting this claim has
been strengthened since AR5 \citep{IPCC2013}. Recent studies, however,
have produced contrasting findings, with some suggesting that drought
has not exhibited a significant trend over the past forty years
\citep{Vicente-Serrano2022, Kogan2020}. \citet{Vicente-Serrano2022}
analyzed the trend in meteorological drought on a global scale, finding
that only in a few regions an increase in the severity of drought was
observed. Moreover, they attributed this increase solely to an increase
in atmospheric evaporative demand (AED) due to higher temperatures,
which in turn enhances vegetation water demand, with important
implications for agricultural and ecological droughts. Also, they state
that \emph{``the increase in hydrological droughts has been primarily
observed in regions with high water demand and land cover change, led by
an increase in agricultural land''}. Similarly, \citet{Kogan2020}
analyzed the drought trend using remotely-sensed vegetation health
indicators, finding that for the globe and main grain-producing
countries, drought has not expanded or intensified during the past 38
years. Nonetheless, \citet{IPCC2021} suggests that there is a medium to
high degree of confidence that rising temperatures will increase the
extent, frequency, and severity of agricultural and ecological droughts.
Also, AR6 \citep{IPCC2023} predicts that many regions of the world will
experience more severe agricultural and ecological droughts even if
global warming stabilizes at 1.5°--2°C. To better evaluate the impact of
drought trends on ecosystems, assessments that correlate meteorological
and soil moisture variables to their effects on vegetation are much
needed.

From 1960 to 2019, land use change has impacted around one-third of the
Earth's surface, which is four times more than previously thought
\citep{Winkler2021}. Multiple studies aim to analyze and forecast
changes in land cover globally \citep{Winkler2021, Song2018} and
regionally
\citep{Chamling2020, Homer2020, Yang2021, Schulz2010, Echeverria2012}.
Some seek to analyze the impact of land cover change on climate
conditions such as temperature and precipitation
\citep{Luyssaert2014, Pitman2012}. There is less research on drought and
its relation to land cover change and vegetation productivity
\citep{Chen2022, Akinyemi2021, Peng2017}. \citet{Peng2017} utilized net
primary productivity to examine the spatial and temporal variations in
vegetation productivity at global level and assess to what extent
drought influenced this variability by comparing the twelve-month
Standardized Precipitation Evapotranspiration Index (SPEI) and land
cover change. According to their findings, drought is responsible for
37\% of the decline and accounts for 55\% of the variability in
vegetation productivity. \citet{Chen2022} instead found poor
correlations (r\textless0.2) between the vegetation productivity trends
against meteorological drought (SPEI of twelve months in December) and
soil moisture at the global level. These studies mostly looked at how
changes in land cover and vegetation productivity are related to a
single drought index (SPEI) obtained for 12 month periods. SPEI takes
into account the combined effect of precipitation and AED as a water
balance, but it does not allow to know the contribution of each variable
on its own. To better understand these contributions on land cover
change and vegetation productivity the following questions may be asked:
i) how do land cover and vegetation productivity respond to short- to
long-term meteorological and soil moisture droughts? And ii) How is this
response different between humid and arid climatic zones? Likewise,
there is a lack of understanding of how the alteration in water supply
and demand is affecting land cover transformations.

To address the previous questions over extensive regions, we can utilize
gridded data on water availability, vegetation conditions, and the
respective drought indices. For monitoring drought, the World
Meteorological Organization recommends the SPI (Standardized
Precipitation Index) \citep{WMO2012}. The SPI is a multi-scalar drought
index that only uses precipitation to assess short- to long-term
droughts. \citet{Vicente-Serrano2010} proposed the Standardized
Precipitation Evapotranspiration Index (SPEI), which incorporates the
temperature effect by subtracting AED from precipitation. SPEI allows
for analyzing the combined effect of precipitation and AED. Since its
formulation, it has been used worldwide for the study and monitoring of
drought \citep{Gebrechorkos2023, Liu2024}. Recently, there has been more
interest in using AED to track droughts separately to better disengage
the effects of precipitation from temperature-dependent effects
\citep{Vicente-Serrano2020}. One of the reasons is that AED is linked to
more flash droughts in limited water regions \citep{Noguera2022}.
\citet{Hobbins2016} and \citet{McEvoy2016} developed the Evaporative
Demand Drought Index (EDDI) to monitor droughts solely using the AED,
and it has proven effective in monitoring flash droughts
\citep{Li2024, Ford2023}. For soil moisture, several drought indices
exist, such as the Soil Moisture Deficit Index (SDMI)
\citep{Narasimhan2005} and the Soil Moisture Agricultural Drought Index
(SMADI) \citep{Souza2021}. \citet{Hao2013} and \citet{AghaKouchak2014}
proposed the Standardized Soil Moisture Index (SSI), which has a similar
formulation as the SPI, SPEI, and EDDI. Thus, many drought indices exist
that allow for a comprehensive assessment of drought on short- to
long-term scales and that allow for the use of single variables from
Earth's water balance (e.g., precipitation, AED, soil moisture).
Climatic variability impacts vegetation development, with unfavorable
conditions such as low precipitation and high temperatures usually
promote a decrease in plant productivity. To monitor the response of
vegetation for large areas, the common practice is to use satellite
data. For example, the Normalized Difference Vegetation Index (NDVI)
derived from frequent satellite observations of red and near infrared
spectral reflectance, has been widely used as a proxy for biomass
production \citep{Camps-Valls2021, Paruelo2016, Helman2014}. For Chile's
cultivated land, \citet{Zambrano2018} used the zcNDVI for assessing
seasonal biomass production in response to drought. Comparing the
various meteo-related and vegetation-based drought indices, we can
further our understanding of the impact of drought on ecosystems.

Chile's diverse climatic and ecosystem types
\citep{Beck2023, Luebert2022} make it an ideal natural laboratory for
studying climate and ecosystems. Additionally, the country has
experienced severe drought conditions that have had significant effects
on vegetation and water storage. North-central Chile has faced a
persistent precipitation deficit since 2010, defined as a mega-drought
\citep{Garreaud2017}, which has impacted the Chilean ecosystem and
consequently makes it highly vulnerable to climate change
\citep{Barria2021, AlvarezGarreton2021}. This mega-drought was defined
by the annual time series of the Standardized Precipitation Index (SPI)
at a time scale of twelve months at the end of each year (December) when
having values below one standard deviation. Some studies have addressed
how this drought affects single ecosystems in terms of forest growth
\citep{Miranda2020, Venegas2018}, forest fire occurrence
\citep{UrrutiaJalabert2018}, and crop productivity
\citep{Zambrano2023, Zambrano2018, Zambrano2016}. The term
``mega-drought'' is used in Chile to describe a prolonged water shortage
that lasts for several years, resulting in a permanent deficit that
impacts the hydrological system \citep{Boisier2018}. Therefore, it is
crucial to evaluate temporal scales that consider the cumulative impact
over a period of several years. In Chile, the relationship between
drought and the environment remains poorly understood. Hence, we aim to
contribute to understanding how climatic and soil moisture droughts
influence ecosystem dynamics in order to provide useful information that
helps for a better understanding of ecological droughts and, at the same
time, helps to make well-informed decisions on adaptation strategies.

@@

\hypertarget{study-area}{%
\section{Study area}\label{study-area}}

Continental Chile has diverse climate conditions with strong gradients
from north to south and east to west \citep{Aceituno2021}
(Figure~\ref{fig-studyArea}a), which determines its great ecosystem
diversity \citep{Luebert2022} (Figure~\ref{fig-studyArea}c). The Andes
Mountains are a main factor in climate variation \citep{Garreaud2009}.
For an aggregated overview of the results of the study, we used the five
Chilean macrozones: ``Norte Grande,'' ``Norte Chico,'' ``Centro,''
``Sur,'' and ``Asutral''. ``Norte Grande'' (17°34'--25°42'S) and ``Norte
Chico'' (25°42'-32°8'S) predominate in an arid desert climate with hot
(Bwh) and cold (Bwk) temperatures. At the south of ``Norte Chico,'' the
climate changes to an arid steppe with cold temperatures (Bsk). In these
two northern regions, the land is mostly bare, with a small surface of
vegetation types such as shrubland and grassland. In the macrozones
``Centro'' (32°08'-36°12'S) and the northern half of ``Sur,''
(36°12'-43°48'S) the main climate is Mediterranean, with warm to hot
summers (Csa and Csb). Land cover in ``Centro'' comprises a significant
amount of shrubland and savanna (50\%), grassland (16\%), forest (8\%),
and croplands (5\%). An oceanic climate (Cfb) predominates in the south
of ``Sur'' and the north of ``Austral'' (43°48'-56°00'S) Those zones
have a large areal extent of forest and grassland. The southern part of
the country has a tundra climate, and in ``Austral,'' it is a cold
semi-arid area with an extended surface of grassland, forest, and, to a
lesser extent, savanna.

@@

{\centering \includegraphics{../output/figs/map_study_con_landcover.png}

@@

\end{figure*}

\hypertarget{materials-and-methods}{%
\section{Materials and Methods}\label{materials-and-methods}}

\hypertarget{data}{%
\subsection{Data}\label{data}}

\hypertarget{gridded-meteorological-and-vegetation-data}{%
\subsubsection{Gridded meteorological and vegetation
data}\label{gridded-meteorological-and-vegetation-data}}

>>>>

\usepackage{lineno}\linenumbers \usepackage{multirow} \usepackage{lscape} \newcommand{\blandscape}{\begin{landscape}} \newcommand{\elandscape}{\end{landscape}}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Journal Name}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The effects of drought on land cover change and vegetation productivity in continental Chile},
  pdfauthor={Francisco Zambrano; Anton Vrieling; Francisco Meza; Iongel Duran-Llacer; Francisco Fernández; Alejandro Venegas-González; Nicolas Raab; Dylan Craven},
  pdfkeywords={drought, land cover change, vegetation
productivity, ecosystem},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\author[8,9]{Francisco Fernández%
%
}
\author[10]{Alejandro Venegas-González%
%
}

@@

\author[12,13]{Dylan Craven%
%
}

\affiliation[1]{organization={Hémera Centro de Observación de la Tierra,
Facultad de Ciencias, Escuela de Ingeniería en Medio Ambiente y
Sustentabilidad, Universidad Mayor},city={Santiago},country={7500994,
Chile.}}
\affiliation[2]{organization={Observatorio de Sequía para la Agricultura
y la Biodiversidad de Chile (ODES), Universidad
Mayor},city={Santiago},country={7500994,
Chile.}}
\affiliation[3]{organization={Faculty of Geo-Information Science and
Earth, University of Twente},city={Enschede},country={The
Netherlands.}}
\affiliation[4]{organization={Facultad de Agronomía y Sistemas
Naturales, Pontificia Universidad Católica de
Chile.},city={Santiago},country={Chile.}}
\affiliation[5]{organization={Instituto para el Desarrollo Sustentable.
Pontificia Universidad Católica de
Chile},city={Santiago},country={Chile.}}
\affiliation[6]{organization={Centro Interdisciplinario de Cambio
Global, Pontificia Universidad Católica de
Chile},city={Santiago},country={Chile.}}
\affiliation[7]{organization={Hémera Centro de Observación de la Tierra,
Facultad de Ciencias, Universidad
Mayor,},city={Santiago},country={7500994,
Chile.}}
\affiliation[8]{organization={Center of Economics for Sustainable
Development (CEDES), Faculty of Economics and Government, Universidad
San
Sebastian},city={Santiago},country={Chile.}}
\affiliation[9]{organization={Center of Applied Ecology and
Sustainability
(CAPES)},city={Santiago},country={Chile.}}
\affiliation[10]{organization={Instituto de Ciencias Agroalimentarias,
Animales y Ambientales (ICA3), Universidad de O'Higgins},city={San
Fernando},country={Chile.}}
\affiliation[11]{organization={, GEMA Center for Genomics, Ecology \&
Environment, Universidad Mayor, Camino La Pirámide Huechuraba
5750},city={Santiago},country={Chile.}}
\affiliation[12]{organization={Data Observatory
Foundation},city={Santiago},country={Chile.}}

@@

\begin{abstract}
Chile has experienced a persistent decrease in water supply, which
impacts the hydrological system and vegetation development. This
persistent period of water scarcity has been defined as a mega-drought.
There is yet insufficient understanding of ecological drought in Chile
due to the limited studies on the relationship between drought and
ecosystem changes. The aim of our study is to evaluate the interaction
of drought, land cover change, and vegetation productivity over
continental Chile. To assess drought, we used drought indices for
atmospheric evaporative demand (AED), water supply, and soil moisture
from short- (1, 3, 6 months) to long-term (12, 24, 36 months) time
scales. We derived the drought indices using monthly ERA5-Land
reanalysis data from 1981 to 2023. We used Moderate-Resolution Imaging
Spectroradiometer (MODIS) datasets to derive information on annual land
cover and monthly vegetation productivity. Our results showed that,
except for the Austral part, Chile has a temporal decreasing trend in
water supply, and across the whole country, there is an increase in AED.
These trends become stronger over longer time scales. We found a
negative trend in vegetation productivity in the north-central area,
which is more prominent for shrubland and savanna as compared to
croplands and forests. The anomaly in soil moisture over the past 12
months (SSI-12) is the most important variable explaining these changes,
followed by anomalies in accumulated precipitation over one to two years
(SPI-12 and SPI-24). The variable importance obtained by random forest
models indicates that drought explains about 12--41\% of the change in
land cover surface across Chile for forest, grassland, shrubland, and
savanna but has little relation to the changes in croplands. The
increase in AED is the main variable associated with the change in land
cover, followed by a reduction in precipitation and soil moisture. Our
findings provide insightful information that could assist in developing
adaptation measures for Chilean ecosystems to cope with climate change
and drought.
\end{abstract}

\begin{keyword}
    drought \sep land cover change \sep vegetation productivity \sep 
    ecosystem
\end{keyword}
\end{frontmatter}

\captionsetup{justification=raggedright,singlelinecheck=false}

\section{Introduction}\label{introduction}

Drought can be classified as 1) meteorological, when precipitation in a
specific period remains below the mean precipitation experienced in the
same period during multiple years (more than 30 years usually); 2)
hydrological, when these anomalies last for long periods (months to
years) and affect water systems; and 3) agricultural, when the deficit
negatively impacts plant health and leads to decreased productivity of
crops or pastures \citep{Wilhite1985}. However, because drought is also
influenced by human activities, \citet{Loon2016} and
\citet{AghaKouchak2021} expanded the drought definition for the
Anthropocene, indicating that the feedback of human decisions and
activities should also be considered (i.e., anthropogenic drought).
Droughts can lead to increased tree mortality \citep{Cheng2024} and
induce alterations in land cover and land use, ultimately affecting
ecosystems \citep{Crausbay2017}. Even though many ecological studies
have at times mistakenly considered ``dry'' conditions as ``drought''
\citep{Slette2019}, ecological drought can be defined as \emph{``an
episodic deficit in water availability that drives ecosystems beyond
thresholds of vulnerability, impacts ecosystem services, and triggers
feedback in natural and/or human systems''} \citep{Crausbay2017}. In
light of current global warming, it is crucial to study the interaction
between drought and ecosystems in order to understand their feedback and
impact on future water security \citep{Bakker2012}.

Global warming, as a result of human-induced greenhouse gas emissions,
has increased the frequency and intensity of drought, according to the
sixth assessment report (AR6) of the Intergovernmental Panel on Climate
Change (IPCC) \citep{IPCC2023}. The evidence supporting this claim has
been strengthened since AR5 \citep{IPCC2013}. Recent studies, however,
have produced contrasting findings, with some suggesting that drought
has not exhibited a significant trend over the past forty years
\citep{Vicente-Serrano2022, Kogan2020}. \citet{Vicente-Serrano2022}
analyzed the trend in meteorological drought on a global scale, finding
that only in a few regions an increase in the severity of drought was
observed. Moreover, they attributed this increase solely to an increase
in atmospheric evaporative demand (AED) due to higher temperatures,
which in turn enhances vegetation water demand, with important
implications for agricultural and ecological droughts. Also, they state
that \emph{``the increase in hydrological droughts has been primarily
observed in regions with high water demand and land cover change, led by
an increase in agricultural land''}. Similarly, \citet{Kogan2020}
analyzed the drought trend using remotely-sensed vegetation health
indicators, finding that for the globe and main grain-producing
countries, drought has not expanded or intensified during the past 38
years. Nonetheless, \citet{IPCC2021} suggests that there is a medium to
high degree of confidence that rising temperatures will increase the
extent, frequency, and severity of agricultural and ecological droughts.
Also, AR6 \citep{IPCC2023} predicts that many regions of the world will
experience more severe agricultural and ecological droughts even if
global warming stabilizes at 1.5°--2°C. To better evaluate the impact of
drought trends on ecosystems, assessments of the relationship between
meteorological and soil moisture variables and their effects on
vegetation are much needed.

From 1960 to 2019, land use change has impacted around one-third of the
Earth's surface, which is four times more than previously thought
\citep{Winkler2021}. Multiple studies aim to analyze and forecast
changes in land cover globally \citep{Winkler2021, Song2018} and
regionally
\citep{Chamling2020, Homer2020, Yang2021, Schulz2010, Echeverria2012}.
Some seek to analyze the impact of land cover change on climate
conditions such as temperature and precipitation
\citep{Luyssaert2014, Pitman2012}. There is less research on drought and
its relation to land cover change and vegetation productivity
\citep{Chen2022, Akinyemi2021, Peng2017}. \citet{Peng2017} utilized net
primary productivity to examine the spatial and temporal variations in
vegetation productivity at global level and assess to what extent
drought influenced this variability by comparing the twelve-month
Standardized Precipitation Evapotranspiration Index (SPEI) and land
cover change. According to their findings, drought is responsible for
37\% of the decline and accounts for 55\% of the variability in
vegetation productivity. \citet{Chen2022} instead found poor
correlations (r\textless0.2) between the vegetation productivity trends
against meteorological drought (SPEI of twelve months in December) and
soil moisture at the global level. These studies mostly looked at how
changes in land cover and vegetation productivity are related to a
single drought index (SPEI) obtained for 12 month periods. SPEI takes
into account the combined effect of precipitation and AED as a water
balance, but it does not allow to know the contribution of each variable
on its own. To better understand these contributions on land cover
change and vegetation productivity the following questions may be asked:
i) how do land cover and vegetation productivity respond to short- to
long-term meteorological and soil moisture droughts? And ii) how is this
response different between humid and arid climatic zones? Likewise,
there is a lack of understanding of how the alteration in water supply
and demand is affecting land cover transformations.

To address the previous questions over extensive regions, we can utilize
gridded data on water availability, vegetation conditions, and the
respective drought indices. For monitoring drought, the World
Meteorological Organization recommends the SPI (Standardized
Precipitation Index) \citep{WMO2012}. The SPI is a multi-scalar drought
index that only uses precipitation to assess short- to long-term
droughts. \citet{Vicente-Serrano2010} proposed the Standardized
Precipitation Evapotranspiration Index (SPEI), which incorporates the
temperature effect by subtracting AED from precipitation. SPEI allows
for analyzing the combined effect of precipitation and AED. Since its
formulation, it has been used worldwide for the study and monitoring of
drought \citep{Gebrechorkos2023, Liu2024}. Recently, there has been more
interest in using AED to track droughts separately to better disentangle
precipitation from temperature-dependent effects
\citep{Vicente-Serrano2020}. One of the reasons is that AED is more
linked to flash droughts in water-limited regions \citep{Noguera2022}.
\citet{Hobbins2016} and \citet{McEvoy2016} developed the Evaporative
Demand Drought Index (EDDI) to monitor droughts solely using the AED,
and it has proven effective in monitoring flash droughts
\citep{Li2024, Ford2023}. For soil moisture, several drought indices
exist, such as the Soil Moisture Deficit Index (SDMI)
\citep{Narasimhan2005} and the Soil Moisture Agricultural Drought Index
(SMADI) \citep{Souza2021}. \citet{Hao2013} and \citet{AghaKouchak2014}
proposed the Standardized Soil Moisture Index (SSI), which has a similar
formulation as the SPI, SPEI, and EDDI. Thus, many drought indices exist
that allow for a comprehensive assessment of drought on short- to
long-term scales and that allow for the use of single variables from
Earth's water balance (e.g., precipitation, AED, soil moisture).
Climatic variability impacts vegetation development, with unfavorable
conditions such as low precipitation and high temperatures usually
promoting a decrease in plant productivity. To monitor the response of
vegetation for large areas, the common practice is to use satellite
data. For example, the Normalized Difference Vegetation Index (NDVI)
derived from frequent satellite observations of red and near infrared
spectral reflectance, has been widely used as a proxy for biomass
production
\citet{Camps-Valls2021};\citet{Paruelo2016};\citet{Helman2014}{]}. For
Chile's cultivated land, \citet{Zambrano2018} used the zcNDVI for
assessing seasonal biomass production in response to drought. Comparing
the various meteo-related and vegetation-based drought indices, we can
further our understanding of the impact of drought on ecosystems.

Chile's diverse climatic and ecosystem types
\citep{Beck2023, Luebert2022} make it an ideal natural laboratory for
studying climate and ecosystems. Additionally, the country has
experienced severe drought conditions that have had significant effects
on vegetation and water storage. North-central Chile has faced a
persistent precipitation deficit since 2010, defined as a mega-drought
\citep{Garreaud2017}, which has impacted the Chilean ecosystem and
consequently makes it highly vulnerable to climate change
\citep{Barria2021, AlvarezGarreton2021}. This mega-drought was defined
by the annual time series of the Standardized Precipitation Index (SPI)
at a time scale of twelve months at the end of each year (December) when
having values below one standard deviation. Some studies have addressed
how this drought affects single ecosystems in terms of forest growth
\citep{Miranda2020, Venegas2018}, forest fire occurrence
\citep{UrrutiaJalabert2018}, and crop productivity
\citep[\citet{Zambrano2016}]{Zambrano2023, Zambrano2018}. The term
``mega-drought'' is used in Chile to describe a prolonged water shortage
that lasts for several years, resulting in a permanent deficit that
impacts the hydrological system \citep{Boisier2018}. Therefore, it is
crucial to evaluate temporal scales that consider the cumulative impact
over a period of several years. In Chile, the relationship between
drought and the environment remains poorly understood. Hence, we aim to
contribute to understanding how climatic and soil moisture droughts
influence ecosystem dynamics in order to provide useful information that
helps for a better understanding of ecological droughts and, at the same
time, helps to make well-informed decisions on adaptation strategies.

@@

\section{Study area}\label{study-area}

Continental Chile has diverse climate conditions with strong gradients
from north to south and east to west \citep{Aceituno2021}
Figure~\ref{fig-studyArea}a, which determines its great ecosystem
diversity \citep{Luebert2022} (Figure~\ref{fig-studyArea}c). The Andes
Mountains are a main factor in climate variation \citep{Garreaud2009}.
For an aggregated overview of the results of the study, we used the five
Chilean macrozones: ``Norte Grande'' (17°34'--25°42'S), ``Norte Chico''
(25°42'-32°8'S), ``Centro'' (32°08'-36°12'S), ``Sur'' (36°12'-43°48'S),
and ``Austral'' (43°48'-56°00'S). ``Norte Grande'' and ``Norte Chico''
predominate in an arid desert climate with hot (Bwh) and cold (Bwk)
temperatures. At the south of ``Norte Chico'', the climate changes to an
arid steppe with cold temperatures (Bsk). In these two northern regions,
the land is mostly bare, with a small surface of vegetation types such
as shrubland and grassland. In the macrozones ``Centro'' and the
northern half of ``Sur'', the main climate is Mediterranean, with warm
to hot summers (Csa and Csb). Land cover in ``Centro'' comprises a
significant amount of shrubland and savanna (50\%), grassland (16\%),
forest (8\%), and croplands (5\%). An oceanic climate (Cfb) predominates
in the south of ``Sur'' and the north of ``Austral''. Those zones have a
large area of forest and grassland. The southern part of the country has
a tundra climate, while ``Austral'' is a cold semi-arid area with an
extended surface of grassland, forest, and, to a lesser extent, savanna.

@@

\centering{

\includegraphics{../output/figs/map_study_con_landcover.png}

@@

\end{figure*}%

\section{Materials and Methods}\label{materials-and-methods}

\subsection{Data}\label{data}

\subsubsection{Gridded meteorological and vegetation
data}\label{gridded-meteorological-and-vegetation-data}

To analyze land cover change, we used the classification scheme by the
IGBP (International Geosphere-Biosphere Programme) from the product
MCD12Q1 Collection 6.1 from MODIS. The MCD12Q1 product is produced for
each year from 2001 to 2022 and defines 17 classes (see Table S1). To
maintain our focus on a large scale and follow the FAO classification
\citep{FAO2022}, we considered native and planted forests as
``forests'', which represent ecosystems dominated by larger trees. To
derive a proxy for vegetation productivity, we used the Normalized
Difference Vegetation Index (NDVI) from the product MOD13A3 Collection
6.1 from MODIS \citep{Didan2015}. MOD13A3 provides vegetation indices
with 1km spatial resolution and monthly frequency. The NASA EOSDIS Land
Processes Distributed Active Archive Center (LP DAAC), USGS Earth
Resources Observation and Science (EROS) Center, Sioux Falls, South
Dakota, provided the MOD13A3 and MCD12Q1 from the online Data Pool,
accessible at \url{https://lpdaa.usgs.gov/tools/data-pool/}.

\begin{table}

\centering{

[!ht]
